---
title: "Xenium Data Analysis Report"
author: ""
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
      toc_depth: 3
    number_sections: true
    fig_caption: yes
    theme: cerulean
    code_folding: hide
editor_options: 
  chunk_output_type: console
params:
  Xenium_output: './'
  lower_threshold: 0.05
  higher_threshold: 0.95
  cell_stats_path: NULL
  metadata_sheet_path: NULL
  sample_name: NULL
  segmentations: NULL
---

```{r setup, include=FALSE}

suppressPackageStartupMessages({
library(Seurat)
library(ggplot2)
library(clustree)
library(dplyr)
#BiocManager::install("RBioFormats")
#library(RBioFormats)
#install.packages('magick',repo = 'https://cran.rstudio.com')
#library(magick)
})

```

# Read in samples

If the data is a raw Count Matrix, meaning that the empty droplets are not filtered out by cellranger pipeline or Drop-Seq pipeline, the emptyDrops algorithm from DropUtils will be run in order to remove the empty droplets.

```{r read in samples,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

if(!is.null(params$segmentations) && !is.na(params$segmentations) && params$segmentations != 'none'){
  Data=Seurat::LoadXenium(params$Xenium_output, segmentations = params$segmentations)
} else {
  Data=Seurat::LoadXenium(params$Xenium_output)
}


metadata=read.csv(paste0(params$Xenium_output,'/cells.csv.gz'),header=T,row.names=1)

## Read and process metadata sheet
metadata_sheet <- if (!is.null(params$metadata_sheet_path) && !is.na(params$metadata_sheet_path)) {
  tryCatch({
    sheet <- read.delim(params$metadata_sheet_path, sep="\t", header=TRUE)
    message("Successfully read metadata sheet")
    
    if (!is.null(params$sample_name) && !is.na(params$sample_name)) {
      sheet <- sheet[sheet$sample_name == params$sample_name, ]
      if (nrow(sheet) == 0) warning(sprintf("No rows found for sample name: %s", params$sample_name))
    }
    sheet
  }, error = function(e) {
    warning(sprintf("Error reading metadata sheet: %s", e$message))
    NULL
  })
} else {
  message("No metadata sheet path provided")
  NULL
}

# Process cell stats if both metadata sheet and stats path exist
combined_stats <- if (!is.null(metadata_sheet) && !is.null(params$cell_stats_path) && !is.na(params$cell_stats_path)) {
  stats_list <- list()
  metadata_cols <- c("condition", "replicate", "tissue")
  
  for (i in 1:nrow(metadata_sheet)) {
    cell_stats_file <- metadata_sheet$cell_stats_path[i]
    if (is.null(cell_stats_file) || is.na(cell_stats_file) || cell_stats_file == "") next
    
    full_path <- file.path(params$cell_stats_path, basename(cell_stats_file))
    if (!file.exists(full_path)) stop(sprintf("Cell stats file not found: %s", full_path))
    
    # Debug: Check file structure before reading
    message(sprintf("Reading cell stats file: %s", full_path))
    
    # Skip first two lines and read header from third line
    lines <- readLines(full_path)
    header_line <- strsplit(lines[3], ",")[[1]]
    message("Header from line 3:")
    message(header_line)
    
    # Count columns in header vs data
    header_cols <- length(header_line)
    if (length(lines) > 3) {
      data_cols <- length(strsplit(lines[4], ",")[[1]])
      message(sprintf("Header columns: %d, Data columns: %d", header_cols, data_cols))
    }
    
    # Try reading with different options, skipping first two lines
    tryCatch({
      stats_data <- read.csv(full_path, skip = 2, check.names = FALSE)
    }, error = function(e) {
      message(sprintf("Error with default read.csv: %s", e$message))
      # Try with different separators
      tryCatch({
        stats_data <<- read.csv(full_path, skip = 2, sep = "\t", check.names = FALSE)
        message("Successfully read with tab separator")
      }, error = function(e2) {
        message(sprintf("Error with tab separator: %s", e2$message))
        # Try reading with fill = TRUE to handle missing columns
        tryCatch({
          stats_data <<- read.csv(full_path, skip = 2, fill = TRUE, check.names = FALSE)
          message("Successfully read with fill = TRUE")
        }, error = function(e3) {
          message(sprintf("Error with fill = TRUE: %s", e3$message))
          # Try reading without header
          tryCatch({
            stats_data <<- read.csv(full_path, skip = 2, header = FALSE, check.names = FALSE)
            message("Successfully read without header")
          }, error = function(e4) {
            # Last resort: read as text and manually parse
            message("Attempting manual parsing...")
            lines <- readLines(full_path)
            if (length(lines) > 3) {
              # Use the third line as header, pad with generic names if needed
              header_line <- strsplit(lines[3], ",")[[1]]
              max_cols <- max(sapply(lines[4:length(lines)], function(x) length(strsplit(x, ",")[[1]])))
              
              if (length(header_line) < max_cols) {
                # Pad header with generic column names
                additional_cols <- paste0("V", (length(header_line) + 1):max_cols)
                header_line <- c(header_line, additional_cols)
                message(sprintf("Padded header with %d additional column names", length(additional_cols)))
              }
              
              # Create a temporary file with corrected header and data
              temp_file <- tempfile(fileext = ".csv")
              writeLines(c(paste(header_line, collapse = ","), lines[4:length(lines)]), temp_file)
              stats_data <<- read.csv(temp_file, check.names = FALSE)
              unlink(temp_file)
              message("Successfully read with manual header correction")
            } else {
              stop("File appears to be empty or has fewer than 3 lines")
            }
          })
        })
      })
    })
    for (col in metadata_cols) {
      if (!is.null(metadata_sheet[[col]][i]) && !is.na(metadata_sheet[[col]][i]) && metadata_sheet[[col]][i] != "") {
        stats_data[[col]] <- metadata_sheet[[col]][i]
      }
    }
    stats_list[[i]] <- stats_data
  }
  do.call(rbind, stats_list)
} else NULL

# Merge metadata with combined stats if they exist
if (!is.null(combined_stats)) {
  # Debug: Check dimensions
  message(sprintf("Metadata dimensions: %d rows, %d cols", nrow(metadata), ncol(metadata)))
  message(sprintf("Combined stats dimensions: %d rows, %d cols", nrow(combined_stats), ncol(combined_stats)))
  message("Combined stats column names:")
  message(paste(colnames(combined_stats), collapse = ", "))
  
  # Add cell_id columns
  metadata$cell_id <- rownames(metadata)
  
  # Check if combined_stats has a cell identifier column and standardize it
  if ("Cell ID" %in% colnames(combined_stats)) {
    # Rename 'Cell ID' to 'cell_id' for consistency
    colnames(combined_stats)[colnames(combined_stats) == "Cell ID"] <- "cell_id"
    message("Renamed 'Cell ID' column to 'cell_id' in combined_stats")
  } else if (!"cell_id" %in% colnames(combined_stats)) {
    # If no cell_id column exists, we can't merge without cell identifiers
    warning("No cell identifier column found in combined_stats - cannot merge")
    updated_metadata <- metadata
  } else {
    # combined_stats already has cell_id, use it as is
    message("Using existing cell_id column from combined_stats")
  }
  
  # Proceed with merge if combined_stats has a cell_id column
  if ("cell_id" %in% colnames(combined_stats)) {
    # Check for overlap between cell IDs
    metadata_cell_ids <- metadata$cell_id
    stats_cell_ids <- combined_stats$cell_id
    overlap <- intersect(metadata_cell_ids, stats_cell_ids)
    
    message(sprintf("Cell ID overlap: %d cells in metadata, %d cells in stats, %d overlapping", 
                   length(metadata_cell_ids), length(stats_cell_ids), length(overlap)))
    
    if (length(overlap) > 0) {
      # Merge the datasets - this will keep all metadata rows and add stats where available
      updated_metadata <- merge(metadata, combined_stats, by="cell_id", all.x=TRUE)
      
      # Ensure the order matches the original metadata
      updated_metadata <- updated_metadata[match(metadata$cell_id, updated_metadata$cell_id), ]
      rownames(updated_metadata) <- updated_metadata$cell_id
      updated_metadata$cell_id <- NULL
      
      message(sprintf("Successfully merged data: %d cells with stats, %d cells without stats", 
                     sum(!is.na(updated_metadata[, ncol(updated_metadata)])), 
                     sum(is.na(updated_metadata[, ncol(updated_metadata)]))))
    } else {
      warning("No overlapping cell IDs found between metadata and combined_stats - using metadata only")
      updated_metadata <- metadata
    }
  } else {
    updated_metadata <- metadata
  }
} else {
  updated_metadata <- metadata
}

Data <- AddMetaData(Data, updated_metadata)

Data=subset(Data,subset=nCount_Xenium>0&nFeature_Xenium>0)
Data=subset(Data,subset=nCount_Xenium>as.numeric(quantile(Data$nCount_Xenium,params$lower_threshold))&
              nCount_Xenium<as.numeric(quantile(Data$nCount_Xenium,params$higher_threshold))&
              nFeature_Xenium>as.numeric(quantile(Data$nFeature_Xenium,params$lower_threshold))&
              nFeature_Xenium<as.numeric(quantile(Data$nFeature_Xenium,params$higher_threshold)))

#Data@misc$background.image=background

```

# QC {.tabset}

In this tab the correlation between number of genes, and number of UMIs and area of cells will be visualized

## Number of Genes per Cells

```{r Number of Genes per Cells, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

VlnPlot(Data,'nFeature_Xenium',pt.size = 0)

```

## Number of UMIs per Cells

```{r Number of UMIs per Cells, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

VlnPlot(Data,'nCount_Xenium',pt.size = 0)

p3 = FeatureScatter(Data, 'nFeature_Xenium','nCount_Xenium',raster = T)

```

## Correlation between number of Genes and UMIs

```{r Correlation between number of Genes and UMIs, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

FeatureScatter(Data, 'nFeature_Xenium','nCount_Xenium',raster = T)

```

# Analysis

This tab will do the normalization, dimension reduction and parameter optmization.

## PCA and selecting number of PCs for UMAP and Clustering

```{r PCA, fig.width=5,fig.height=5,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

Data=SCTransform(Data,assay = 'Xenium')

Data=RunPCA(Data,npcs = 100,assay = 'SCT')

candidates=abs(diff(Data@reductions$pca@stdev))

candidates=which(candidates>mean(candidates))

candidates=max(candidates)

plot(Data@reductions$pca@stdev)+abline(v=(candidates+1))


```

## Clustering {.tabset}

```{r Clustering, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

Data=RunUMAP(Data,dims = 1:(candidates+1))

Data=FindNeighbors(Data,dims = 1:(candidates+1))

Data=FindClusters(Data,resolution = seq(2.0,0.1,-0.1))


names=paste0(DefaultAssay(Data),'_snn_res.')
SC3_Stability=clustree(Data,prefix = names)
SC3_Stability.results=SC3_Stability$data
SC3_Stability.results=SC3_Stability.results[,c(names,'sc3_stability')]
colnames(SC3_Stability.results)[1]='resolution'
SC3_Stability.results.mean=aggregate(sc3_stability~resolution,SC3_Stability.results,mean)
colnames(SC3_Stability.results.mean)[2]='sc3_stability_mean'

candidates=abs(diff(SC3_Stability.results.mean$sc3_stability_mean))
candidates=which(candidates>mean(candidates))

Idents(Data)=paste0(DefaultAssay(Data),'_snn_res.',SC3_Stability.results.mean$resolution[max(candidates)])

Data$seurat_clusters=Idents(Data)

Data=BuildNicheAssay(Data,fov = 'fov',group.by = 'seurat_clusters')

Data$niches=as.factor(Data$niches)

DefaultAssay(Data)='Xenium'


Data=NormalizeData(Data,assay='Xenium')


Data.markers=FindAllMarkers(Data,assay='Xenium',only.pos = T)

saveRDS(Data,"Output.rds")
```

### UMAP of Clusters

```{r UMAP of Clusters, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

DimPlot(Data)

```

### UMAP of Niches

```{r UMAP of Niches, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

DimPlot(Data,group.by='niches')

```

### Spatial plot of clusters

```{r Spatial plot of clusters, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

ImageDimPlot(Data,cols = "polychrome",dark.background = T,group.by='seurat_clusters')

```

### Spatial plot of Niches

```{r Spatial plot of Niches, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

ImageDimPlot(Data,cols = "polychrome",dark.background = T,group.by='niches')

```

## Cluster Markers

```{r Top gene markers for clusters,error=FALSE,warning=FALSE,message=FALSE,echo=FALSE}

top10 = Data.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)

DT::datatable(top10,filter = "top",options = list(autoWidth = TRUE))

```