---
title: "Xenium Data Analysis Report"
author: "Via Scientific"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
      toc_depth: 3
    number_sections: true
    fig_caption: yes
    theme: cerulean
    code_folding: hide
editor_options: 
  chunk_output_type: console
params:
  Xenium_output: './'
  lower_threshold: 0.05
  higher_threshold: 0.95
---

```{r setup, include=FALSE}

suppressPackageStartupMessages({
library(Seurat)
library(ggplot2)
library(clustree)
library(dplyr)
#BiocManager::install("RBioFormats")
#library(RBioFormats)
#install.packages('magick',repo = 'https://cran.rstudio.com')
#library(magick)
})

```

# Read in samples

If the data is a raw Count Matrix, meaning that the empty droplets are not filtered out by cellranger pipeline or Drop-Seq pipeline, the emptyDrops algorithm from DropUtils will be run in order to remove the empty droplets.

```{r read in samples,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

Data=Seurat::LoadXenium(params$Xenium_output)

metadata=read.csv(paste0(params$Xenium_output,'/cells.csv.gz'),header=T,row.names=1)

Data=AddMetaData(Data,metadata)

Data=subset(Data,subset=nCount_Xenium>0&nFeature_Xenium>0)
Data=subset(Data,subset=nCount_Xenium>as.numeric(quantile(Data$nCount_Xenium,params$lower_threshold))&
              nCount_Xenium<as.numeric(quantile(Data$nCount_Xenium,params$higher_threshold))&
              nFeature_Xenium>as.numeric(quantile(Data$nFeature_Xenium,params$lower_threshold))&
              nFeature_Xenium<as.numeric(quantile(Data$nFeature_Xenium,params$higher_threshold)))

#Data@misc$background.image=background

```

# QC {.tabset}

In this tab the correlation between number of genes, and number of UMIs and area of cells will be visualized

## Number of Genes per Cells

```{r Number of Genes per Cells, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

VlnPlot(Data,'nFeature_Xenium',pt.size = 0)

```

## Number of UMIs per Cells

```{r Number of UMIs per Cells, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

VlnPlot(Data,'nCount_Xenium',pt.size = 0)

p3 = FeatureScatter(Data, 'nFeature_Xenium','nCount_Xenium',raster = T)

```

## Correlation between number of Genes and UMIs

```{r Correlation between number of Genes and UMIs, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

FeatureScatter(Data, 'nFeature_Xenium','nCount_Xenium',raster = T)

```

# Analysis

This tab will do the normalization, dimension reduction and parameter optmization.

## PCA and selecting number of PCs for UMAP and Clustering

```{r PCA, fig.width=5,fig.height=5,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

Data=SCTransform(Data,assay = 'Xenium')

Data=RunPCA(Data,npcs = 100,assay = 'SCT')

candidates=abs(diff(Data@reductions$pca@stdev))

candidates=which(candidates>mean(candidates))

candidates=max(candidates)

plot(Data@reductions$pca@stdev)+abline(v=(candidates+1))


```

## Clustering {.tabset}

```{r Clustering, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

Data=RunUMAP(Data,dims = 1:(candidates+1),)

Data=FindNeighbors(Data,dims = 1:(candidates+1))

Data=FindClusters(Data,resolution = seq(2.0,0.1,-0.1))


names=paste0(DefaultAssay(Data),'_snn_res.')
SC3_Stability=clustree(Data,prefix = names)
SC3_Stability.results=SC3_Stability$data
SC3_Stability.results=SC3_Stability.results[,c(names,'sc3_stability')]
colnames(SC3_Stability.results)[1]='resolution'
SC3_Stability.results.mean=aggregate(sc3_stability~resolution,SC3_Stability.results,mean)
colnames(SC3_Stability.results.mean)[2]='sc3_stability_mean'

candidates=abs(diff(SC3_Stability.results.mean$sc3_stability_mean))
candidates=which(candidates>mean(candidates))

Idents(Data)=paste0(DefaultAssay(Data),'_snn_res.',SC3_Stability.results.mean$resolution[max(candidates)])

Data$seurat_clusters=Idents(Data)

Data=BuildNicheAssay(Data,fov = 'fov',group.by = 'seurat_clusters')

Data$niches=as.factor(Data$niches)

DefaultAssay(Data)='Xenium'


Data=NormalizeData(Data,assay='Xenium')


Data.markers=FindAllMarkers(Data,assay='Xenium',only.pos = T)

saveRDS(Data,"Output.rds")
```

### UMAP of Clusters

```{r UMAP of Clusters, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

DimPlot(Data)

```

### UMAP of Niches

```{r UMAP of Niches, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

DimPlot(Data,group.by='niches')

```

### Spatial plot of clusters

```{r Spatial plot of clusters, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

ImageDimPlot(Data,cols = "polychrome",dark.background = T,group.by='seurat_clusters')

```

### Spatial plot of Niches

```{r Spatial plot of Niches, fig.width=10,fig.height=10,error=FALSE,warning=FALSE,message=FALSE,results=FALSE,echo=FALSE}

ImageDimPlot(Data,cols = "polychrome",dark.background = T,group.by='niches')

```

## Cluster Markers

```{r Top gene markers for clusters,error=FALSE,warning=FALSE,message=FALSE,echo=FALSE}

top10 = Data.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)

DT::datatable(top10,filter = "top",options = list(autoWidth = TRUE))

```
